Magic-free Code Extraction of Generic Programs and Recursion Schemes in Coq

### 2024-03-18 paper/paper.tex
Revision of catamorphisms

  This paper presents a formalisation of hylomorphisms (a recursion scheme
  capturing divide-and-conquer algorithms) in Coq, along with their algebraic
  laws. The key contribution of this paper is that this formalisation is fully
  axiom-free, and it allows for the extraction of safe, idiomatic OCaml
  code for a variety of recursive programs. We formalise a series of divide-and
  conquer, dynamic programming, and mutual recursion programs and demonstrate
  that the extracted OCaml code for the programs formalised in our framework is
  efficient, resembles code that a human programmer would write, and contains
  no occurrences of \ocaml{Obj.magic}. We also present a machine-checked proof
  of the well-known short-cut fusion optimisation.
\end{abstract}

@@

Not many approaches mechanise structured recursion
schemes in proof assistants.  Recently, Abreu et al.~\cite{AbreuDHJMS23} encode
an algebraic approach to divide-and-conquer computations, in which termination
is entirely enforced by the typing discipline of the recursion scheme. However,
although their approach solves the problem of termination proofs, as well as
the performance of the code that is run \emph{within Coq}, it no longer allows
the extraction of idiomatic OCaml code.  By this, we refer to code
that (1) does not preserve the recursive structure of common implementations;
and (2) lead to unsafe casts \ocaml{Obj.magic} in the generated code. 
In OCaml, \ocaml{Obj.magic} allows for unsafe casts between types.
Although the use of \ocaml{Obj.magic} should be safe in the code
extracted by Coq, it has been proven that, for higher-order programs, simple
interoperations can lead to incorrect behaviour or even
segfaults~\cite{forster:hal-04329663}. Therefore, the avoidance of axioms and
\ocaml{Obj.magic} means that the extracted code will have stronger
safety guarantees, and will allow the safe interoperation with other OCaml
code. The absence of \ocaml{Obj.magic} provides even stronger guarantees to the
extracted code. In particular, since the code that we extract from Coq is
implemented in a total language, we know that there can be no $\bot$ in its
definition. This allows the programmer to use ``fast-and-loose'' reasoning on
the extracted code~\cite{DanielssonHJG06}, which would otherwise be unsafe if
the code contained \ocaml{Obj.magic}. For example, the programmer would need to
manually check that the inputs satisfy the necessary assumptions which would be
impossible to enforce in the presence of unsafe casts. Overall, the absence of
\ocaml{Obj.magic} makes it easer to reason about, and interoperate with
extracted code.

This work presents the first formalisation in Coq of \emph{hylomorphisms}, as
well as their algebraic laws that: (1) is \emph{axiom-free}; and (2)
allows the extraction of idiomatic OCaml code. While programmers
still need to reason about the termination of their programs, our mechanisation
will allow the use of the algebraic laws of hylomorphisms for program
reasoning, as well as extracting to idiomatic code. As an
example our mechanisation of \ocaml{qsort} as a hylomorphism will produce
the following code:
\begin{coqcode}
let rec qsort = function | [] -> []
  | h :: t -> let (l, r) = partition (fun x0 -> (<=) x0 h) t in
              let cx = fun p -> match p with | Lbranch -> l | Rbranch -> r in 
              app (qsort (cx Lbranch)) (h :: (qsort (cx Rbranch)))
\end{coqcode}
The contributions we make in this paper are as follows:
\begin{itemize}
  \item We provide a framework for \emph{generic programming} with recursion
    schemes in Coq, with a proof of their algebraic laws and 
    program equivalences, that allows clean code extraction.
  \item We formalise divide and conquer, dynamic programming, and
    mutual recursion algorithms.
  \item We verified the short-cut fusion optimisation in Coq and
    extract optimised code to OCaml.
\end{itemize}

@@

\subsection{Algebras and Catamorphisms}
\label{sec:algebras} An algebra is a set \coq{X} (the \emph{carrier} of the
algebra) together with some operations on it\footnote[1]{Sometimes and algebra
comes equipped with an equational theory, e.g. the theory of monoids, but here
we focus on algebraically free algebras}.  For example, given such a type
\coq{X} we can define the monoid operations as \coq{u : unit -> X} for unit and
\coq{m : X * X -> X} for multiplication. Notice that \coq{F X = unit + X * X}
is a functor on \coq{X} which means we can equivalently describe these two maps
as a single one of type \coq{F X -> X} which we call an algebra for the functor
\coq{F} satisfying the unit and associative laws of the monoid.  In this work
we will be using algebras to describe the operations of a data type and so we
will not be needing additional equations on these operations.

Since in this work we will be considering only polynomial functors we need to
formalise algebras in terms of containers and containers extensions. Given a
type \coq{A} and a container \coq{F} an algebra for \coq{F} over the carrier
\coq{A} is a map of the following type:
\begin{coqcode}
Notation Alg F A := (App F A ~> A).
\end{coqcode}

\dcas{The writing below is confusing, but I am not sure how to fix it ... Can't
we start with the list-generating functor? The monoid example is great, but as 
soon as we move to talking about least-fixed points, it feels as if we'd better
go straight into F X = unit + A * X}
Now, the \emph{easiest} and most effective way to turn any type $A$ into a
monoid is to define it as the least solution to the equation
\begin{coqcode}
Free A = A + unit + Free A * Free A
\end{coqcode}
with \coq{Free : Type -> Type}. In Coq, this is the least fixed-point of the
functor \coq{unit + A + X * X} on \coq{X}.  This is isomorphic to the type of
lists over \coq{A} hence it is the least fixed-point for the
functor \coq{unit + A * X} as well. This is the least ``set'' generated by the
constructors \coq{Empty : unit -> List A} and \coq{Cons : A * List A -> List A}.
These constructors give an algebra \coq{LFix_in : unit + A * List A -> List A}
which has and inverse which we call \coq{LFix_out}.

We now formalise this idea using container extensions.  For a container
extension a data type is formalised as an inductive data type parametrised by
the functor represented by \coq{F}:
\begin{coqcode}
Inductive LFix `(F : Cont Sh P) : Type := LFix_in { LFix_out : App F (LFix F) }.
\end{coqcode}
% To generalise recursion schemes we abstract the shape of the data using functors.
% Assuming a base functor \coq{F} a catamorphism is then the unique
% solution to the equation
% \begin{minted}{coq}
% cata alg = alg . fmap (cata alg) . InOp
% \end{minted}
% which does precisely what we described in the previous paragraph.
%
% where $F$ is a \emph{base functor}, $Ff$ is the \emph{functorial action} which
% can be thought of the program $\operator{fmap}\; f$, $a$ is a program of type
% $F A \to A$ (an $F$-algebra),  $c$ is a program of type $C \to FC$ (a
% $F$-coalgebra), $\InOp$ is the isomoprhism on the least fixed-point of $F$,
% namely $\InOp : \mu F \to F\mu F$, and $\OutOp$ is the isomorphism on the
% greatest fixed-point of $F$, namely $F \nu F \to \nu F$. When the unique
% solutions to the above equations exist we denote them as $\cata{a}$ and
% $\ana{c}$ respectively.
%
%%\dcas{Boring stuff}
%The definition of \coq{LFix} in terms of \coq{App} requires us to define our
%own custom induction scheme for initial algebras:
%\begin{coqcode}
%Lemma LFix_ind [P : LFix -> Prop]
%  : (forall x : App F LFix,
%        (forall e : Pos (shape x), P (cont x e)) ->
%        P (LFix_in x))
%    -> forall l : LFix, P l.
%\end{coqcode}
We define \coq{LFix} as a setoid, where its equivalence relation can be
described as the least fixed point of \coq{AppR}.  We define smart constructors
for the isomorphism of least fixed points as respectful morphisms.
\begin{coqcode}
l_in : App F (LFix F) ~> LFix F                l_out : LFix F ~> App F (LFix F)
\end{coqcode}

The most simple and well-known recursion scheme is the \emph{fold} or
\emph{catamorphism} which consumes data. First a fold \emph{destructs} a data
type using \coq{LFix_out}, then calls itself recursively and then folds the
results of the recursive call using an \coq{F}-algebra.
\begin{coqcode}
  Definition cata_f `{F : Cont Sh P} (alg : Alg F A) : LFix -> A
:= fix f (x : LFix) := match x with
  | LFix_in ax => let (sx, kx) := ax in alg (MkApp sx (fun e => f (kx e))) end.
\end{coqcode}
It is easy to show that this function is a respectful morphism of
\coq{F}-algebras. In fact, it is possible to define it as a map of the
following type:
\begin{coqcode}
  cata : forall `{F : Cont Sh P} `{setoid A}, Alg F A ~> LFix ~> A
\end{coqcode}
We prove that catamorphisms satisfy the universal law, that is that there is a
unique morphism of \coq{F}-algebras from the least fixed-point of \coq{F}:
\begin{coqcode}
Lemma cata_univ `{eA : setoid A} (alg : Alg F A) (f : LFix ~> A)
    : f =e cata alg <-> f =e alg \o fmap f \o l_out.
\end{coqcode}

@@

\section{Related Work}
%
%\begin{itemize}
%  \item McBride devised a generalisation of Bove and Capretta's general
%recursive function monad~\cite{McBride15, BoveC01}
%  \item Guarded recursion for productive definitions and general recursion
%~\cite{AtkeyM13, PaviottiMB15}
%  \item PACO library~\cite{HurNDV13}
%  \item Sized Types \cite{HughesPS96, abs-1202-3496}
%  \item Copatterns \cite{AbelPTS13, AbelP16}
%  \item Not that related: extraction for mutable arrays~\cite{SAKAGUCHI2020102372}
%        \item Extraction of Distributed Mobile code in Coq~\cite{MiculanP12}.
%\end{itemize}
Encoding recursion schemes in Coq is not new. We compare our work with other
encodings of program calculation techniques in Coq (Section~\ref{sec:calccoq}),
recursion schemes in Coq (Sections~\ref{sec:calccoq} and~\ref{sec:reccoq}).
Finally, we compare our encoding of recursion schemes to other forms of
termination checking (Section~\ref{sec:termcheck}, and sized types
(Section~\ref{sec:sizedT}), as a way to guarantee termination of nonstructural
recursion.

\subsection{Program Calculation}\label{sec:calccoq}

Within the domain of program optimization, program calculation serves as a
well-established programming technique aimed at deriving efficient programs
from their naive counterparts through systematic program
transformation~\cite{gibbonsSquiggol}. This area has been extensively explored
over the years. Tesson et al.\ demonstrated the efficacy of leveraging Coq to
establish an approach for implementing a robust system dedicated to verifying
the correctness of program transformations for functions that manipulate
lists~\cite{10.1007/978-3-642-17796-5_10}. Murata and Emoto went further and
formalised recursion schemes in Coq~\cite{MurataE19}. Their development does
not include hylomorphisms and dynamorphisms, and relies on the functional
extensionality axiom, as well as further extensionality axioms for each
coinductive datatype that they use. They do not discuss the extracted OCaml
code from their formalisation.  Larchey-Wendling and Monin encode recursion
schemes in Coq, by formalising computational graphs of
algorithms~\cite{larchey2022braga}. Their work does not focus on encoding
generic recursion schemes, and proving their algebraic laws.  Castro-Perez et
al.~\cite{farmsCastro} encode the laws of hylomorphisms as part of the type
system of a functional language to calculate parallel programs from
specifications. Their work focuses on parallelism, and they do not formalise
their approach in a proof assistant, and the laws of hylomorphisms are axioms
in their system.

@@

\subsection{Termination Checking}\label{sec:termcheck}

@@

\subsubsection{Divide and Conquer Recursion}\label{sec:reccoq}

\subsubsection{Sized Types}\label{sec:sizedT}

@@

>>>>

  This paper presents a formalisation of hylomorphisms (a recursion scheme
  capturing divide-and-conquer algorithms) in Coq, along with their algebraic
  laws. The key contribution of this paper is that this formalisation is fully
  axiom-free, and it allows for the extraction of safe, clean, idiomatic OCaml
  code for a variety of recursive programs. We formalise a series of divide-and
  conquer, dynamic programming, and mutual recursion programs and demonstrate
  that the extracted OCaml code for the programs formalised in our framework is
  efficient, resembles code that a human programmer would write, and contains
  no occurrences of \ocaml{Obj.magic}. We also present a machine-checked proof
  of the well-known short-cut fusion optimisation.
\end{abstract}

@@

Not many approaches mechanise structured recursion
schemes in proof assistants.  Recently, Abreu et al.~\cite{AbreuDHJMS23} encode
an algebraic approach to divide-and-conquer computations, in which termination
is entirely enforced by the typing discipline of the recursion scheme. However,
although their approach solves the problem of termination proofs, as well as
the performance of the code that is run \emph{within Coq}, it no longer allows
the extraction of clean, naturally-looking code.  By this, we refer to code
that (1) does not preserve the recursive structure of common implementations;
and (2) lead to unsafe casts \ocaml{Obj.magic} in the generated code. 
In OCaml, \ocaml{Obj.magic} allows for unsafe casts between types.
Although the use of \ocaml{Obj.magic} should be safe in the code
extracted by Coq, it has been proven that, for higher-order programs, simple
interoperations can lead to incorrect behaviour or even
segfaults~\cite{forster:hal-04329663}. Therefore, the avoidance of axioms and
\ocaml{Obj.magic} means that the extracted code will have stronger
safety guarantees, and will allow the safe interoperation with other OCaml
code.

This work presents the first formalisation in Coq of \emph{hylomorphisms}, as
well as their algebraic laws that: (1) is \emph{axiom-free}; and (2)
allows the extraction of naturally-looking, clean OCaml code. While programmers
still need to reason about the termination of their programs, our mechanisation
will allow the use of the algebraic laws of hylomorphisms for program
reasoning, as well as extracting to clean, naturally-looking code. As an
example our mechanisation of \ocaml{qsort} as a hylomorphism will produce
the following code:
\begin{coqcode}
let rec qsort = function | [] -> []
  | h :: t -> let (l, r) = partition (fun x0 -> (<=) x0 h) t in
              let cx = fun p -> match p with | Lbranch -> l | Rbranch -> r in 
              app (qsort (cx Lbranch)) (h :: (qsort (cx Rbranch)))
\end{coqcode}
The contributions we make in this paper are as follows:
\begin{itemize}
  \item We provide a framework for \emph{generic programming} with recursion
    schemes in Coq, with a proof of their algebraic laws and 
    program equivalences, that allows clean code extraction.
  \item We formalise divide and conquer, dynamic programming, and
    mutual recursion algorithms.
  \item We verified the short-cut fusion optimisation in Coq and
    extract optimised code to OCaml.
\end{itemize}

@@

\subsection{Algebras and Catamorphisms}
\label{sec:algebras} An algebra is a set \coq{A} (the
\emph{carrier} of the algebra) together with some operations on it sometimes
subject to some equational laws. For example, given such a type \coq{A} we can
define the monoid operations as \coq{u : unit -> A} for unit
and \coq{m : A * A -> A} for multiplication. Notice
that \coq{F X = unit + X * X} is a functor on \coq{X} which means we can
equivalently describe these two maps as a single one of type \coq{F X -> X}
which we call the algebra for the functor \coq{F} satisfying the unit and
associative laws of the monoid. In this work we will be using algebras to
describe the operations of a data type and so we will not be needing additional
equations on these operations.

Given a type \coq{A} and a functor \coq{F}, an \coq{F}-algebra is a pair given
by a type \coq{A} called the \emph{carrier} of a \emph{structure map} of
type \coq{App F A ~> A}:
\begin{coqcode}
Notation Alg F A := (App F A ~> A).
\end{coqcode}
The least fixed-point for a functor \coq{F} is an instance of an \coq{F}-algebra
where the structure map is an isomorphism. This is sometimes referred to as the
\emph{initial algebra} for a functor \coq{F}. We will explain the reason behind this name shortly. \footnote[1]{To be more precise,
initial algebras are isomorphisms by Lambek's lemma, but isomorphisms of
functors do not necessarily correspond to initial algebras.}  We define the
least fixed-point of \coq{F} as an inductive type:
\begin{coqcode}
Inductive LFix `(F : Cont Sh P) : Type := LFix_in { LFix_out : App F (LFix F) }.
\end{coqcode}
\mpav{David, make it more Coq syntax like}
where \coq{LFix_in} is the \coq{F}-algebra while \coq{LFix_out} is its inverse.
As an example, the initial \coq{F}-algebra for the
functor \coq{F X = unit + A * X} is the type of lists with the \coq{F}-algebra
being defined by the empty list \coq{Empty : unit -> LFix F} and the cons
operation \coq{Cons : A * LFix F -> LFix F}.

We define \coq{LFix} as a setoid, where its equivalence relation can be
described as the least fixed point of \coq{AppR} and we define smart
constructors for the isomorphism of least fixed points as respectful morphisms:
\begin{coqcode}
l_in : App F (LFix F) ~> LFix F                l_out : LFix F ~> App F (LFix F)
\end{coqcode}

The least fixed-point for \coq{F} is the initial \coq{F}-algebra in the sense
that it gives rise to an inductive recursion scheme.  Specifically, for any
other \coq{F}-algebra there exists a \emph{unique} map, known as a \emph{fold} or
\emph{catamorphism}, such that it structurally deconstructs the data type
using \coq{LFix_out}, calls itself recursively and then composes the result of
the recursive call using the given an \coq{F}-algebra. In other words, for any
give \coq{F}-algebra there exists a unique \coq{F}-algebra homomorphism from the
initial one. We define it in Coq as follows:
\begin{coqcode}
  Definition cata_f `{F : Cont Sh P} (alg : Alg F A) : LFix -> A
:= fix f (x : LFix) := match x with
  | LFix_in ax => let (sx, kx) := ax in alg (MkApp sx (fun e => f (kx e))) end.
\end{coqcode}
It is easy to show that this function is a respectful morphism of
\coq{F}-algebras. In fact, it is possible to define it as a map of the
following type:
\begin{coqcode}
  cata : forall `{F : Cont Sh P} `{setoid A}, Alg F A ~> LFix ~> A
\end{coqcode}
We prove that catamorphisms satisfy the universality property we explained previously:
\begin{coqcode}
Lemma cata_univ `{eA : setoid A} (alg : Alg F A) (f : LFix ~> A)
    : f =e cata alg <-> f =e alg \o fmap f \o l_out.
\end{coqcode}
In other words, if there is any other \coq{f} with the same structural recursive
shape as the catamorphism on the algebra \coq{alg} then it must be equal to that
catamorphism.

@@

\section{Related Work}
%
%\begin{itemize}
%  \item McBride devised a generalisation of Bove and Capretta's general
%recursive function monad~\cite{McBride15, BoveC01}
%  \item Guarded recursion for productive definitions and general recursion
%~\cite{AtkeyM13, PaviottiMB15}
%  \item PACO library~\cite{HurNDV13}
%  \item Sized Types \cite{HughesPS96, abs-1202-3496}
%  \item Copatterns \cite{AbelPTS13, AbelP16}
%  \item Not that related: extraction for mutable arrays~\cite{SAKAGUCHI2020102372}
%        \item Extraction of Distributed Mobile code in Coq~\cite{MiculanP12}.
%\end{itemize}

\subsection{Program Calculation}

Within the domain of program optimization, program calculation serves as a
well-established programming technique aimed at deriving efficient programs
from their naive counterparts through systematic program
transformation~\cite{gibbonsSquiggol}. This area has been extensively explored
over the years. Tesson et al.\ demonstrated the efficacy of leveraging Coq to
establish an approach for implementing a robust system dedicated to verifying
the correctness of program transformations for functions that manipulate
lists~\cite{10.1007/978-3-642-17796-5_10}. Murata and Emoto went further and
formalised recursion schemes in Coq~\cite{MurataE19}. Their development does
not include hylomorphisms and dynamorphisms, and relies on the functional
extensionality axiom, as well as further extensionality axioms for each
coinductive datatype that they use. They do not discuss the extracted OCaml
code from their formalisation.  Larchey-Wendling and Monin encode recursion
schemes in Coq, by formalising computational graphs of
algorithms~\cite{larchey2022braga}. Their work does not focus on encoding
generic recursion schemes, and proving their algebraic laws.  Castro et
al.~\cite{farmsCastro} encode the laws of hylomorphisms as part of the type
system of a functional language to calculate parallel programs from
specifications. Their work focuses on parallelism, and they do not formalise
their approach in a proof assistant, and the laws of hylomorphisms are axioms
in their system.

@@

\subsection{Termination Checking}

@@

\subsubsection{Sized Types}

@@

\subsubsection{Divide and Conquer Recursion}
